---
title: Missing Data Patterns
author: Haziq Jamil
page-layout: full
grid:
  sidebar-width: 0px
editor_options: 
  chunk_output_type: console
html-table-processing: none
---

## Load data 

```{r}
#| echo: true
#| eval: true 
#| file: data.R
```

## Missing data patterns

Let's take a look at the missing data patterns in the housing data.
For simplicity, we focus on the year 2024 and houses in the Brunei-Muara district.

```{r}
hsp24 <- 
  hsp |>
  filter(year(date) == 2024, district == "Brunei Muara") |>
  select(price, mukim, plot_area, floor_area, beds, baths)
glimpse(hsp24)
```

There are a total of `r nrow(hsp24)` houses in the Brunei-Muara district in 2024.
There are a couple of missing values in the data set, `r sum(is.na(hsp24))` to be exact.
We can visualise this using the `{mice}` and `{ggmice}` package.

```{r}
library(mice)
library(ggmice)
```

Let's take a look at the missing data patterns in the housing data.

```{r}
plot_pattern(hsp24, square = !TRUE, rotate = !TRUE) 
```

We first note that in this instance, there are no missing values for price and spatial location.
However, there is a pattern in the missing data, so the missingness mechanism is not *completely* at random. 
It seems to be the case that the missing data is *at random* (MAR), which means that the missing data is dependent on the observed data.
We can test this using a logistic regression.

```{r}
hsp24 |>
  mutate(across(everything(), is.na)) |>
  glm(formula = beds ~ plot_area + floor_area + baths, family = binomial) |>
  summary()
```

What we can see is that the `beds` variable missingness is dependent on the `bath` missingness, which means that if the number of bathrooms is missing, the number of bedrooms is also likely to be missing (odds ratio increases by $e^{8.52} = `r round(exp(8.52), 1)`$).

On the other hand, the missingness of the `floor_area` variable does not seem to be dependent on the other variables, which means that this variable is missing *completely* at random (MCAR).

```{r}
hsp24 |>
  mutate(across(everything(), is.na)) |>
  glm(formula = floor_area ~ plot_area + beds + baths, family = binomial) |>
  summary()
```

Let's visualise the missing points as a scatter plot.

```{r}
ggmice(hsp24, aes(y = price / 1000, x = floor_area)) +
  geom_point() +
  geom_smooth(method = "lm", col = "black") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(y = "Price (BND 1,000)", x = "Floor area (sq. ft.)")
```


## Simulating missingness mechanism

Suppose we have a full data set and we want to simulate a missingness mechanism.
For this purpose, we have the following data sets:

1. DATA SET A: The original data set containing missing values.
2. DATA SET B: The "complete cases" data set, i.e. subsetting DATA SET A containing no missing values whatsoever.

Suppose we want to induce missingness in B with the same pattern as A.
Here's the code to do it:

```{r}
set.seed(123)
datA <- hsp24
datB <- drop_na(hsp24)

# Step 1: Calculate the proportion of missing values for each variable in the original data
pat <-
  mice::md.pattern(datA, plot = FALSE) |>
  as.data.frame() |>
  rownames_to_column(var = "freq") |>
  select(-V7) |>
  slice_head(n = -1) |>
  unite("pattern", beds:floor_area, sep = "", remove = FALSE) |>
  mutate(
    freq = as.numeric(freq),
    prop = freq / sum(freq),
    id = row_number()
  ) |> 
  as_tibble()
pat

# Step 2: Randomly sample from this multinomial distribution
mis <- sample(x = pat$pattern, size = nrow(datB), prob = pat$prop, replace = TRUE)
# Check that it matches the original pattern
round(prop.table(table(mis)), 3)
rev(round(pat$prop, 3))

# Step 3: Introduce missing values into the complete dataset based on these proportions
for (i in 1:nrow(datB)) {
  na_flags <- as.integer(strsplit(mis[i], "")[[1]]) == 0L
  datB[, c("beds", "baths", "plot_area", "floor_area")][i, na_flags] <- NA
}
```

Let's visualise the missing data patterns in the simulated data set.

```{r}
plot_pattern(datB, square = !TRUE, rotate = !TRUE) 
```

Some patterns are there, but some are not. This is expected when some of the missing patterns are rare.